\textbf{Too Beautiful to be Fake: Attractive Faces are Less Likely to be Judged as Artificially Generated}

### 2023-04-12 submission/manuscript.Rmd

keywords          : "attractiveness, AI-generated images, fiction, fake news, sense of reality"
wordcount         : "4970"
bibliography      : references.bib
floatsintext      : yes
linenumbers       : yes
draft             : no
mask              : no
figurelist        : yes
tablelist         : no
footnotelist      : no
classoption       : "man"
output            : papaja::apa6_pdf
csl: utils/apa.csl
header-includes:
  - \usepackage[labelfont=bf, font={scriptsize, color=gray}]{caption}
editor_options:
  chunk_output_type: console
---

@@

AI-generated content, in particular realistic images, is becoming commonplace and carries important risks for misinformation and black-mailing [@viola2023designed], emphasizing the need to understand the different components that come into play in the formation of reality beliefs.
This exploratory study primarily aims at investigating the effect of facial attractiveness on simulation monitoring, i.e., on the beliefs that an image is real or artificially generated. Based on the *affective reality theory* 
[outlined in @makowski2018cognitive; @makowski2019phenomenal, see **BLOGPOST**  for a summary],
which suggests that salient and emotional stimuli are perceived to be more real (up to a point of reversal after which beliefs of fiction becomes used as an emotion regulation strategy), we hypothesize a quadratic relationship between perceived realness and attractiveness: faces rated as highly attractive or unattractive will more likely be believed to be real. We expect a similar relationship with trustworthiness ratings given its well-established link with attractiveness [@liefooghe2022faces; @bartosik2021you; @garrido2017kdef; @little2011facial], and a positive relationship with familiarity (as more familiar faces would appear as more salient, self-relevant and anchored in reality). Additionally, we will further explore the link shared by dispositional traits, such as personality and attitude towards AI, with simulation monitoring tendencies. This study aims beyond the investigation of the discriminative accuracy between "true" photos and "true" artificially-generated images, focusing on the beliefs that a stimulus is real or fake independently of its true nature.

@@

A related issue is the effect on reality judgement resulting from the explicit instruction in our cover story, which states that "about half of the faces were AI-generated and the other half real photos". Considering this prior information given to participants, our enthusiasm that most people did indeed believe a high number of stimuli to be fake might be unwarranted. After all, in principle, a control condition should not have required a cover story with AI-generated images. However, the distinction real/fake is hard to operationalize and introduce to participants in a vacuum. That said, future studies should study the impact of these higher-order expectations on ratings [for instance, @tucciarelli2020 found that merely mentioning that some faces were AI generated decreased on average the trustworthiness ratings for all faces] as well as on the simulation monitoring process itself (i.e., the "criterion": would people form and distribute judgements differently). This can be studied by modulating this expectation in a controlled fashion (e.g., "most of the images *but a few* are real" vs. "most of the images *but a few* are fake") or inventing some implicit way of measuring reality belief that would not require the explicit introduction of the concept of fake vs. real to participants.

Moreover, although consistent in their directions across  models and variables, the magnitude of the effects found in the study was relatively small, suggesting that the facial features measured in the study were not the key determinants of simulation monitoring. Hence, beyond exploring new potential mechanisms, future studies should include a more thorough debriefing to try to capture what conscious strategies (if any) the participants used (e.g., focusing on some features of the stimulus - like hair or eyes in the case of faces) to guide their reality beliefs.

>>>>

keywords          : "attractiveness, AI-generated images, fiction, fake news, sense of reality"
wordcount         : "5223"
bibliography      : references.bib
floatsintext      : yes
linenumbers       : yes
draft             : no
mask              : no
figurelist        : yes
tablelist         : no
footnotelist      : no
classoption       : "man"
output            : papaja::apa6_pdf
csl: utils/apa.csl
header-includes:
  - \usepackage[labelfont=bf, font={scriptsize, color=gray}]{caption}
editor_options:
  chunk_output_type: console
---

@@

AI-generated content, in particular realistic images, is becoming commonplace and carries important risks for misinformation and black-mailing [@viola2023designed], emphasizing the need to understand the different components that come into play in the formation of reality beliefs.
This exploratory study primarily aims at investigating the effect of facial attractiveness on simulation monitoring, i.e., on the beliefs that an image is real or artificially generated. Based on the *affective reality theory*
[outlined in @makowski2018cognitive; @makowski2019phenomenal, see @makowski2023affectivefor a summary],
which suggests that salient and emotional stimuli are perceived to be more real (up to a point of reversal after which beliefs of fiction becomes used an emotion regulation strategy), we hypothesize a quadratic relationship between perceived realness and attractiveness: faces rated as highly attractive or unattractive will more likely be believed to be real. We expect a similar relationship with trustworthiness ratings given its well-established link with attractiveness [@liefooghe2022faces; @bartosik2021you; @garrido2017kdef; @little2011facial], and a positive relationship with familiarity (as more familiar faces would appear as more salient, self-relevant and anchored in reality). Additionally, we will further explore the link shared by dispositional traits, such as personality and attitude towards AI, with simulation monitoring tendencies. This study aims beyond the investigation of the discriminative accuracy between "true" photos and "true" artificially-generated images, focusing on the beliefs that a stimulus is real or fake, independently of its true nature.

@@

Another issue is the impact on reality judgements of the prior explicit instruction that "about half of the faces were AI-generated and the other half real photos". Given this prior information given to participants, it migjt seems like our enthusiasm pertaining to the finding that most people did indeed believe a high number of stimuli to be fake might be unwarranted: all it shows is that participants followed the instructions. First of all, even if that was the case, the fact that our beliefs of reality can be so easily re-programmed with basic instructions ad lead to high-confidence answers remains an interesting phenomenon. Moreover, it is to note that the paradigm did not instruct participants to balance their answers according to a certain distribution (e.g., 50-50), merely providing a description of the dataset. The fact that no presentation order effect was found on reality beliefs gives credit to the assumption that participants did not try to actively distribute their responses to match the instructions, in which case we would have expected a different pattern, for instance the first items judged as real (the initial "true" belief of the participants), and later a hard bias towards responding "fake" (as participants realize that all stimuli are of similar nature and that they have to "make up" for the prevalence of their "real" answers to fulfil the instructions).

That said, the potential confounding effect of the instructions still exists, and a control condition without the cover story with AI-generated images would in-principle be welcome. However, the distinction real/fake is hard to operationalize and introduce to participants in a vacuum (asking them to discriminate real from fake without introducing the context and meaning of "fake" seems hardly feasible). That said, future studies should study the impact of these higher-order expectations on ratings [for instance, @tucciarelli2020 found that merely mentioning that some faces were AI generated decreased on average the trustworthiness ratings for all faces] as well as on the simulation monitoring process itself (i.e., the "criterion": would people form and distribute judgements differently). This can be studied by modulating this expectation in a controlled fashion (e.g., "most of the images *but a few* are real" vs. "most of the images *but a few* are fake") or inventing some implicit way of measuring reality belief that would not require the explicit introduction of the concept of fake vs. real to participants.

Finally, it is important to note that although consistent in their directions across models and variables, the magnitude of the effects found in the study was relatively small, suggesting that the facial features measured in the study were not the key determinants of simulation monitoring. Hence, beyond exploring new potential mechanisms, future studies should include a more thorough debriefing to try to capture what conscious strategies (if any) the participants used (e.g., focusing on some features of the stimulus - like hair or eyes in the case of faces) to guide their reality beliefs.
